Type theory in type theory with single substitutions

### 2024-09-17 types_2024_postproc/p.tex
stuff

\begin{abstract}
Type theory can be described as a generalised algebraic theory. This
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Algebraic definitions of type theory include Ehrhard's
definition of model, categories with families (CwFs), contextual
categories, Awodey's natural models, C-systems, B-systems. With the
exception of B-systems, these notions are based on a parallel
substitution calculus where substitutions form a category. In this
paper we define a single substitution calculus (SSC) for type theory
and show that the SSC syntax and the CwF syntax are isomorphic for a
theory with dependent function space and a hierarchy of universes. SSC
only includes single substitutions and single weakenings, and eight
equations relating these: four equations describe how to substitute
variables, and there are four equations on types which are needed to
typecheck other equations. All the other equations are substitution
(naturality) rules or computation rules for different type
formers. SSC provides a simple, minimalistic alternative to parallel
substitution calculi or B-systems for defining type theory. SSC
relates to CwF as extensional combinatory calculus relates to lambda
calculus. All the results in this paper were formalised in Agda.
\end{abstract}

@@

TODO: say that we assume the QIIT version of this.

$\alpha$-normal forms could also be called substitution normal forms.

In this section we show that a category with families (CwF) equipped
with a hierarchy of Coquand-universes and $\Pi$ types can be derived
from the single substitution syntax defined in \autoref{sec:tt}.

TODO: show that the SSC operations are derivable from CwF.

We conjecture that the same works for any SOGAT.

\section{stuff}

The following four equations correspond to equations in B-systems which cannot
be derived from the equations of SSC, these are versions of equations in section
\ref{sec:tt} lifted under arbitrary amount of $\blank^+$s.
\begin{equation}\label{eq:lifted}
\begin{aligned}
  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \\
  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \\
  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \\
  B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B
\end{aligned}
\end{equation}
The correspoinding equations on terms are also needed. Note that we do not have
the non-lifted term versions of the last two equations in SSC.

@@

A type or term is $\alpha$-normal if it does not contain instantiation
operations, except at the leaves of the syntax tree as variables. However,
$\alpha$-normal terms can still contain $\beta$/$\eta$ redexes, so that we do
not need to do full normalization. We define variables and $\alpha$-normal forms
as inductive predicates in $\Prop$.
\begin{itemize}
  \item $\q:\Tm\,(\Gamma\ext A)\ (A[\p])$ is a variable.
  \item $x[\p]:\Tm\,(\Gamma\ext A)\ (B[\p])$ is a variable if $x:\Tm\,\Gamma\,B$
  is a variable.
  \item $x:\Tm\,\Gamma\,A$ is $\alpha$-normal if $x$ is a variable.
  \item $\Pi\,A\,B:\Ty\,\Gamma\,i$ is $\alpha$-normal if $A$ and $B$ are
  $\alpha$-normal types.
  \item $\app\,f\,a:\Tm\,\Gamma\,(B[\langle a \rangle])$ is $\alpha$-normal if
  $A$ and $B$ are $\alpha$-normal types, $f:\Tm\,\Gamma\,(\Pi\,A\,B)$ and
  $a:\Tm\,\Gamma\,A$ are $\alpha$-normal terms.
\end{itemize}
The predicate is defined similarly for the other type and term formers. Notably,
we do not state that instantiated types/terms are $\alpha$-normal. It is
important to truncate the $\alpha$-normal predicates to be propositions to
ensure that it preserves the $\beta$/$\eta$ conversion rules.

@@

\begin{lemma}
  \label{thm:alpha}
  The $\alpha$-normal predicate holds for any type and term.
\end{lemma}
\begin{proof}
  By induction on the syntax, $\alpha$-normalizing the substitutions at the same
  time.
\end{proof}

Instead of doing induction on $\alpha$-normal forms for each of the equations in
(\ref{eq:lifted}), we define a general lemma which can lift any equation between
instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
freely added identity and composition operations, we do not require it to
satisfy the laws of a category as we will not compare $\Sub^*$s directly for
equality, only between types and terms instantiated with $\Sub^*$s. All
instantiation operations, lifting operations, and substitution rules can be
redefined for $\Sub^*$.
\begin{lemma}
  \label{thm:lifting}
  Given and $\sigma, \delta : \Sub^*\,\Delta\,\Gamma$, if $A[\sigma] =
  A[\delta]$ and $x[\sigma] = x[\delta]$ for any $A:\Ty\,\Gamma$ and variable
  $x:\Tm\,\Gamma\,A$, then
  \begin{itemize}
    \item $\Omega[\sigma] = \Omega[\delta]$ for $\Omega:\Tel\,\Gamma$
    \item $A[\sigma^{+^\Omega}] = A[\delta^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
    \item $a[\sigma^{+^\Omega}] = a[\delta^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
  \end{itemize}
  Note that the later equations depend the the earlier ones.
\end{lemma}
\begin{proof}
  Assuming the first equation, we prove that $x[\sigma^{+^\Omega}] =
  x[\delta^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
  the variable. Then the two latter equations can be proven by mutual induction
  on $\alpha$-normalized types and terms, still assuming the first equation.
  Finally we prove the first equation by induction on the telescope, using the
  previously proven equation for types, discharging its assumption.
\end{proof}

We can simulate parallel substitutions using $\Sub^*$ as iterated single
substitutions, however it does not satisfy the equations of parallel
substitutions. Thus we define parallel substitutions as a list of terms with a
map into $\Sub^*$, and reuse instantiation operation of $\Sub^*$ to convert it
into a sequence of instantiations of single substitutions.
\begin{alignat*}{10}
  & \Tms && : \Con\ra\Con\ra\Con \\
  & \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
  & \epsilon && : \Tms\,\Gamma\,\diamond \\
  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A) \\
  & \llcorner\epsilon\lrcorner && := \id\circ\p\circ\dots\circ\p \\
  & \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
\end{alignat*}
The computation of instantiating with $\Tms$ is illustrated below.
\begin{gather*}
  B[\llcorner\epsilon,a_1,a_2,\dots,a_{n-1},a_n\lrcorner] = \\
  B[\p^{+^n}]\dots[\p^{+^n}][\langle a_1 \rangle^{+^{n-1}}][\langle a_2 \rangle^{+^{n-2}}]\dots[\langle a_{n-1} \rangle^+][\langle a_n \rangle]
\end{gather*}
All CwF operations and equations can be defined with $\Tms$ by induction.

@@

Finally, we will remove equation $[\p^+][\langle\q\rangle]$ by making
$\Pi\eta$ conditional. This needs another change: we substitute
$\Pi\beta$ for a more general variant which is also conditional on the
same equation.
\begin{alignat*}{10}
  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b
\end{alignat*}
But first we need to check that the usage of $\Pi\beta$ above when
proving $B[\langle a\rangle][\gamma] = B[\gamma^+][\langle
  a[\gamma]\rangle]$ is derivable from $\Pi\beta'$. Assuming a
$\hat{B} : \Tm\,(\Gamma\ext A)\,\U$,
\begin{alignat*}{10}
  & \Pi\beta^\U :{} &&  \lam\,\hat{B}\cdot a && {=}(\q[\langle\rangle]) \\
  & && \lam\,\hat{B}\cdot(\q[\langle a\rangle])\,\, && {=}([\p][\langle\rangle]) \\
  & && ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle])\,\, && {=}(\cdot[]'\,\U[]) \\
  & && ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] && {=}(\Pi\beta'\,\U[]) \\
  & && \hat{B}[\langle a\rangle].
\end{alignat*}
Now we derive the assumption of $\Pi\beta/\eta'$ for any type $B : \Ty\,(\Gamma\ext A)$.
\begin{alignat*}{10}
  & B[\p^+][\langle\q\rangle] && {=}(\U\beta) \\
  & \El\,(\cd\,B)[\p^+][\langle\q\rangle] && {=}(\El[]) \\
  & \El\,(\cd\,B[\p^+][\langle\q\rangle]) && {=}(\Pi\beta^\U) \\
  & \El\,((\lam\,(\cd\,B[\p^+])\cdot\q) && {=}(\lam[]) \\
  & \El\,((\lam\,(\cd\,B)[\p]\cdot\q) && {=}(\Pi\beta'\,\U[]) \\
  & \El\,(\cd\,B) && {=}(\U\beta) \\
  & B
\end{alignat*}
We summarise this section by formally stating what we defined.

@@

\begin{definition}[CwF]\label{def:cwf}
  A category with $\N$-many families is defined as a category (objects
  denoted $\Con$, morphisms $\Sub$) with a terminal object $\diamond$
  ($\epsilon$ denotes the unique morphism into it); for each $i$ a
  presheaf of types (action on objects denoted $\Ty\,\blank\,i$,
  action on morphisms $\blank[\blank]$); for each $i$ a locally
  representable dependent presheaf $\Tm$ over $\Ty\,\blank\,i$
  (actions denoted $\Tm$, $\blank[\blank]$, local representability is
  denoted $\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con$
  with an isomorphism
  $
  (\p\circ\blank,\q[\blank]) : \Sub\,\Delta\,(\Gamma\ext A) \cong (\gamma:\Sub\,\Delta\,\Gamma)\times\Tm\,\Delta\,(A[\gamma]) : (\blank,\blank)
  $
  natural in $\Delta$.
\end{definition}

@@

@@

    & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p][\p^+]\cdot(\delta[\p][\p^+]\cdot\q))\cdot\q)) && {=}(\Pi\beta) \\

@@

@@

In this paper we introduced a new single substitution calculus which
can be seen as a minimalistic variant of B-systems. We illustrate its
usage through the definition of a type theory with dependent function
space and a hierarchy of universes. We aim for accessibility: this
paper can be seen as a tutorial introduction to the syntax of type
theory inside type theory. All the operations and equations are
well-motivated and provide a minimalistic description of the syntax of
type theory. It is minimalistic in the sense that the syntax is
isomorphic to the CwF-syntax, but there are more models.

The relationship of our single substitution calculus and the CwF-based
theory is similar to that of combinatory logic and lambda calculus
\cite{DBLP:conf/fscd/AltenkirchKSV23}. Combinatory logic has more
models than lambda calculus, but the syntaxes are
isomorphic. Similarly, the single substitution calculus has more
models, but the syntaxes are isomorphic.

A more abstract, higher level description of a language is simply
stating that a language is an algebraic theory: the terms of the
language are given by a sort of the theory, the different term formers
are operations, and the conversion relation is specified by the
equations. Such a description contains less information than the
traditional one, for example, conversion is not oriented, so it is not
clear how to turn it into a rewriting relation; we cannot choose how
much annotations we add to terms, they are always fully
annotated. this is similar to how asts contain less information than
strings, for example, the bracket-removal strategy is not encoded in
asts. the algebraic description automatically induces a notion of
model and the existence of an initial model
\cite{DBLP:journals/pacmpl/KaposiKA19} which we call the syntax. The
algebraic approach is especially natural for languages with dependent
types where typing and conversion have to be specified mutually
\cite{DBLP:conf/popl/AltenkirchK16}.

QIIT is nice because syntax and semantics are given together. The
notion of model/semantics is traditionally given separately, and work
has to be done to show that the above specified syntax generates a
model. For example, this was the goal of the initiality project for
homotopy type theory \cite{initiality-agda}. Maybe list work on type theory
using QIITs, see TYPES mailing list.

We can abstract over these strategies and define our language as a
second-order generalised algebraic theory
\cite{DBLP:journals/corr/abs-1904-04097}, but to state initiality of
our syntax, at some point we need to turn it into a first-order
algebraic theory \cite{DBLP:conf/fscd/BocquetKS23}.

We conjecture that any SOGAT can be turned into an SSC.

Separate sort for variables, without the extra type equations, by
using assumptions, and proving everything by induction.

If we have var, then we cannot do termification to get a CwF from a var-SSC.

@@

>>>>

\begin{abstract}
Type theory can be described as a generalised algebraic theory. This
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Algebraic definitions of type theory include Ehrhard's
definition of model, categories with families (CwFs), contextual
categories, Awodey's natural models, C-systems, B-systems. With the
exception of B-systems, these notions are based on a parallel
substitution calculus where substitutions form a category. In this
paper we define a single substitution calculus (SSC) for type theory
and show that the SSC syntax and the CwF syntax are isomorphic for a
theory with dependent function space and a hierarchy of universes. SSC
only includes single substitutions and single weakenings, and eight
equations relating these: four equations describe how to substitute
variables, and there are four equations on types which are needed to
typecheck other equations. All the other equations are substitution
(naturality) rules or computation rules for different type
formers. SSC provides a simple, minimalistic alternative to parallel
substitution calculi or B-systems for defining type theory. SSC
relates to CwF as extensional combinatory calculus relates to lambda
calculus. In case of additional type formers, we show that SSC models
are actually equivalent to CwF-based models in a weak sense. Most
results in this paper were formalised in Agda.
\end{abstract}

@@

In this section we show that our SSC-based syntax is isomorphic to the
CwF-based syntax for the same theory (namely the SOGAT
(\ref{eq:tt})). We conjecture that the method presented here works
generically for any SOGAT.

\begin{definition}[CwF]\label{def:cwf}
  A category with $\N$-many families is defined as a category (objects
  denoted $\Con$, morphisms $\Sub$) with a terminal object $\diamond$
  ($\epsilon$ denotes the unique morphism into it); for each $i$ a
  presheaf of types (action on objects denoted $\Ty\,\blank\,i$,
  action on morphisms $\blank[\blank]$); for each $i$ a locally
  representable dependent presheaf $\Tm$ over $\Ty\,\blank\,i$
  (actions denoted $\Tm$, $\blank[\blank]$, local representability is
  denoted $\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con$
  with an isomorphism
  $
  (\p\circ\blank,\q[\blank]) : \Sub\,\Delta\,(\Gamma\ext A) \cong (\gamma:\Sub\,\Delta\,\Gamma)\times\Tm\,\Delta\,(A[\gamma]) : (\blank,\blank)
  $
  natural in $\Delta$).
\end{definition}

The following four equations correspond to equations in B-systems which cannot
be derived from the equations of SSC, these are versions of equations in Section
\ref{sec:tt} lifted under arbitrary amount of $\blank^+$s.
\begin{equation}\label{eq:lifted}
\begin{aligned}
  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \\
  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \\
  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \\
  B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B
\end{aligned}
\end{equation}
The corresponding equations on terms are also needed. Note that we do not have
the non-lifted term versions of the last two equations in SSC.

@@

A type or term is $\alpha$-normal if it does not contain instantiation
operations, except at the leaves of the syntax tree as variables. However,
$\alpha$-normal terms can still contain $\beta$/$\eta$ redexes, so that we do
not need to do full normalization. We define variables and $\alpha$-normal forms
as inductive predicates in $\Prop$ as follows.
\begin{itemize}
  \item $\q:\Tm\,(\Gamma\ext A)\ (A[\p])$ is a variable.
  \item $x[\p]:\Tm\,(\Gamma\ext A)\ (B[\p])$ is a variable if $x:\Tm\,\Gamma\,B$
  is a variable.
  \item $x:\Tm\,\Gamma\,A$ is $\alpha$-normal if $x$ is a variable.
  \item $\Pi\,A\,B:\Ty\,\Gamma\,i$ is $\alpha$-normal if $A$ and $B$ are
  $\alpha$-normal types.
  \item $\app\,f\,a:\Tm\,\Gamma\,(B[\langle a \rangle])$ is $\alpha$-normal if
  $A$ and $B$ are $\alpha$-normal types, $f:\Tm\,\Gamma\,(\Pi\,A\,B)$ and
  $a:\Tm\,\Gamma\,A$ are $\alpha$-normal terms.
\end{itemize}
The predicate is defined similarly for the other type and term formers. Notably,
we do not state that instantiated types/terms are $\alpha$-normal (except variables). It is
important to truncate the $\alpha$-normal predicates to be propositions to
ensure that it preserves the $\beta$/$\eta$ conversion rules.

@@

\begin{lemma}
  \label{thm:alpha}
  The $\alpha$-normal predicate holds for any type and term.
\end{lemma}
\begin{proof}
  By induction on the syntax, $\alpha$-normalizing the substitutions
  at the same time. Note that induction on the syntax refers to the
  elimination principle of the corresponding QIIT.
\end{proof}

Instead of doing induction on $\alpha$-normal forms for each of the equations in
(\ref{eq:lifted}), we define a general lemma which can lift any equation between
instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
freely added identity and composition operations, we do not require it to
satisfy the laws of a category as we will not compare $\Sub^*$s directly for
equality, only between types and terms instantiated with $\Sub^*$s. All
instantiation operations, lifting operations, and substitution rules can be
redefined for $\Sub^*$.
\begin{lemma}
  \label{thm:lifting}
  Given $\sigma, \delta : \Sub^*\,\Delta\,\Gamma$, if $A[\sigma] =
  A[\delta]$ and $x[\sigma] = x[\delta]$ for any $A:\Ty\,\Gamma$ and variable
  $x:\Tm\,\Gamma\,A$, then
  \begin{itemize}
    \item $\Omega[\sigma] = \Omega[\delta]$ for $\Omega:\Tel\,\Gamma$
    \item $A[\sigma^{+^\Omega}] = A[\delta^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
    \item $a[\sigma^{+^\Omega}] = a[\delta^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
  \end{itemize}
  Note that the later equations depend the the earlier ones.
\end{lemma}
\begin{proof}
  Assuming the first equation, we prove that $x[\sigma^{+^\Omega}] =
  x[\delta^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
  the variable. Then the two latter equations can be proven by mutual induction
  on $\alpha$-normalized types and terms, still assuming the first equation.
  Finally we prove the first equation by induction on the telescope, using the
  previously proven equation for types, discharging its assumption.
\end{proof}

We can simulate parallel substitutions using $\Sub^*$ as iterated single
substitutions, however it does not satisfy the equations of parallel
substitutions. Thus we define parallel substitutions as a list of terms with a
map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to convert it
into a sequence of instantiations of single substitutions.
\begin{alignat*}{10}
  & \Tms && : \Con\ra\Con\ra\Con \\
  & \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
  & \epsilon && : \Tms\,\Gamma\,\diamond \\
  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A) \\
  & \llcorner\epsilon\lrcorner && := \id\circ\p\circ\dots\circ\p \\
  & \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
\end{alignat*}
The computation of instantiating with $\Tms$ is illustrated below.
\begin{gather*}
  B[\llcorner\epsilon,a_1,a_2,\dots,a_{n-1},a_n\lrcorner] = \\
  B[\p^{+^n}]\dots[\p^{+^n}][\langle a_1 \rangle^{+^{n-1}}][\langle a_2 \rangle^{+^{n-2}}]\dots[\langle a_{n-1} \rangle^+][\langle a_n \rangle]
\end{gather*}
All CwF operations and equations can be defined with $\Tms$ by induction.

@@

Finally, we will remove equation $[\p^+][\langle\q\rangle]$ by making
$\Pi\eta$ conditional. This needs another change: we substitute
$\Pi\beta$ for a more general variant which is also conditional on the
same equation.
\begin{alignat*}{10}
  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b
\end{alignat*}
But first we need to check that the usage of $\Pi\beta$ above when
proving $B[\langle a\rangle][\gamma] = B[\gamma^+][\langle
  a[\gamma]\rangle]$ is derivable from $\Pi\beta'$. Assuming a
$\hat{B} : \Tm\,(\Gamma\ext A)\,\U$,
\begin{alignat*}{10}
  & \Pi\beta^\U :{} &&  \lam\,\hat{B}\cdot a && {=}(\q[\langle\rangle]) \\
  & && \lam\,\hat{B}\cdot(\q[\langle a\rangle])\,\, && {=}([\p][\langle\rangle]) \\
  & && ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle])\,\, && {=}(\cdot[]'\,\U[]) \\
  & && ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] && {=}(\Pi\beta'\,\U[]) \\
  & && \hat{B}[\langle a\rangle].
\end{alignat*}
Now we derive the assumption of $\Pi\beta/\eta'$ for any type $B : \Ty\,(\Gamma\ext A)$.
\begin{alignat*}{10}
  & B[\p^+][\langle\q\rangle] && {=}(\U\beta) \\
  & \El\,(\cd\,B)[\p^+][\langle\q\rangle] && {=}(\El[]) \\
  & \El\,(\cd\,B[\p^+][\langle\q\rangle]) && {=}(\Pi\beta^\U) \\
  & \El\,(\lam\,(\cd\,B[\p^+])\cdot\q) && {=}(\lam[]) \\
  & \El\,(\lam\,(\cd\,B)[\p]\cdot\q) && {=}(\Pi\beta'\,\U[]) \\
  & \El\,(\cd\,B) && {=}(\U\beta) \\
  & B
\end{alignat*}
We summarise this section by formally stating what we defined.

@@

@@

  \begin{equation}
  \begin{alignedat}{10}\label{eq:functor}

@@

    & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p][\p^+]\cdot(\delta[\p][\p^+]\cdot\q))\cdot\q)) && {=}(\Pi\beta, {\cdot}[]) \\
    & \lam\,(A[\p]\cdot(\gamma[\p][\p^+][\langle\q\rangle]\cdot(\delta[\p][\p^+][\langle\q\rangle]\cdot\q))) && {=}([\p][^+]) \\
    & \lam\,(A[\p]\cdot(\gamma[\p][\p][\langle\q\rangle]\cdot(\delta[\p][\p][\langle\q\rangle]\cdot\q))) && {=}([\p][\langle\rangle]) \\

@@

  \end{alignedat}
  \end{equation}

@@

Can we remove even more structural rules? the termification doesn't
work if we don't have e.g. $[\p][^+]$ and $[\p][\langle\rangle]$ for all terms, and

Separate sort for variables, without the extra type equations, by
using assumptions, and proving everything by induction. If we have
var, then we cannot do termification to get a CwF from a var-SSC, see
the proof of equation (\ref{eq:functor}).

@@

\phantom{A}
